Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-24 analysis/paper/paper.Rmd
added SAR data, and reference, with pdf. added some preliminary impacts to adult returns to the manuscript

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]; all of which have become depleted in recent decades. Declines in survival of local juvenile Chinook Salmon have been associated with hydrosystem passage [@Schaller2014], ocean conditions and regional climate [@Welch2021; @Gosselin2021], and changes to the freshwater habitat including removal of beavers, river channel simplification, water withdrawals, and the proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*) among others [@IdahoOSCTeam2019]. Moreoever, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

One potentially important, but perhaps under-appreciated source of mortality for Chinook Salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces created by dams on the Columbia and Snake rivers have created favorable slow-water habitat for several species of piscivorous fish known to consume juvenile Chinook Salmon including both non-native (e.g., Smallmouth Bass *Micropterus dolomieu*, Walleye *Sander vitreus*) and native (e.g., Northern Pikeminnow *Ptychocheilus oregonensis*) species. In the Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of non-native Smallmouth Bass during some seasons [@Erhardt2018]. And non-native Walleye are considered a potentially important source for juvenile Chinook Salmon mortality in the Snake River[@Widener2021], particularly as their abundance has significantly increased surrounding Snake River dam facilities between 2008 and 2020 related to their preference for slow water habitats.

In addition to non-native species, Northern Pikeminnow are a native, piscivorous fish preferring warm, slow water habitats. Consequently, they too have benefited from dams on the Columbia River and become abundant predators on salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Predation by piscivorous fishes, including Northern Pikeminnow, may have been a factor in the below-average survival of wild Chinook Salmon in 2020 [@Widener2021]. However, there is a dearth of current data on piscivorous predator populations in the Columbia River.

Habitats containing slower water velocities and other attributes that support piscivorous predators are not limited to dams and reservoirs. One such area is the Deadwater Slough in the Salmon River, Idaho which is an unnaturally slow and deep reach of the mainstem river approximately 1.5 kilometers in length and occupying 25.3 acres (Figure \@ref(fig:map)). The slough contains "reservoir-like" conditions which typically favor piscivorous fish predators like Northern Pikeminnow or Smallmouth Bass [@Watkins2015], and further, lacks hydrological and structural features and fish cover (e.g., large wood or substrate, woody debris, bank undercuts) that can provide essential refuge from predation for juvenile fishes. As a result, predation of juvenile Chinook Salmon and other local ESA-listed salmonids (Sockeye Salmon *O. nerka*, redband steelhead *O. mykiss gairdneri*) has been cited as a concern for populations in the Upper Salmon River

Spring/summer Chinook Salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014]. The DSR migrants leave the natal spawning area as subyearlings between June and november and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal spawning areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migratory tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency.

The Deadwater Slough is in a reach of the Salmon River believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, the slough is part of the migratory pathway for not only juvenile Chinook Salmon (both DSR and NRR emigrants), but also for the endangered Snake River Sockeye Salmon population and several population of threatened Snake River steelhead. As such, the main Salmon River has been the target of recent radio telemetry studies to examine the downstream movement, distribution, and (apparent survival) of juvenile salmonid emigrants. @Axel2015 demonstrated decreased rates of emigration and survival for Sockeye Salmon from Redfish Lake during the spring emigration. Further, recent telemetry studies of Chinook salmon emigration during fall and early winter months indicated a decreased transition probability in the reach spanning Deadwater Slough (approximately 10% less than surrounding reaches).

@@

Text here...

@@

@@

<!--
* Total population estimates of Pike Minnow. Chart or table, but likely a chart.
* What is that equate to as a density? e.g., how many pikeminnow per 10 square meters? Is that a reasonable number?
-->

PSD for northern pikeminnow across all three years was 42.26% or 42.26% of fish were over the quality size classification 400 mm TL we denoted in the Methods, demonstrating a high percentage of above average size classs present within Deadwater Slough.

@@

>>>>

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been associated with hydrosystem passage (@Schaller2014), ocean conditions and regional climate (@Welch2021; @Gosselin2021), and freshwater habitat modification.

All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, river simplification, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

One potentially important, but perhaps under-appreciated source of mortality for Chinook Salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. There are generally two maechanisms that explain these high predation zones. First, migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces created by dams on the Columbia and Snake rivers have created favorable slow-water habitat for several species of piscivorous fish known to consume juvenile Chinook Salmon including Walleye *Sander vitreus*, Smallmouth Bass *Micropterus dolomieu*, and Northern Pikeminnow *Ptychocheilus oregonensis*. In the Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of non-native Smallmouth Bass during some seasons [@Erhardt2018]. Walleye also prefer slow water habitats and are considered a potentially important source for Juvenile Chinook Salmon mortality [@Widener2021], particularly as their abundance has significantly increased in Snake River dam facilities between 2008 and 2020.

In addition to non-native species, Northern Pikeminnow are a native, piscivorous fish preferring slow water habitats. Consequently, they too have benefited from dams on the Columbia River and become abundance predators on salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Predation by piscivorous fish may have been a factor in the below-average survival of wild Chinook Salmon in 2020 [@Widener2021], however there is a dearth of current data on piscivorous predator populations in the Columbia River.

Habitat containing slow-velocities and other attributes that support piscivorous predators are not limited to dams and reservoirs. One such area is the Deadwater Slough in the Salmon River, which occupies a 25.3-acre reach of unnaturally slow and deep water (Figure \@ref(fig:map)). The Deadwater Slough contains little to no cover, increased depth, and fine substrate which favor piscivorous fish predators (e.g., Northern Pikeminnow, Smallmouth Bass) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].

<!--
* @Axel2015 found increased mortality in a particular reach spanning location x to location y in the mainstem Salmon River on a study of juvenile (hatchery?) sockeye salmon released near Redfish Lake and actively migrating downriver.
  - in their case, these were actively migrating juveniles, and so transitional probabilities can likely be interpreted (almost) directly as survival
  - A previous study showed that the lowest transitional probabilities of Red Fish Lake Sockeye occurred in the Dead Water reach.
* @Ackerman2018 and @Porter2019, @Porter2020 found reduced (apparent) survival in a reach spanning near the town of North Fork to just downstream of Deadwater Slough.
  - in these cases, there's a chance that juvenile Chinook could exhibit concealment behavior prior to winter and so transition probability is interpreted as the probability of a fish surviving AND moving
* Anecdotally, locals (and others? maybe NOAA?) informed us that northern pikeminnow and perhaps other species are prevalent in Deadwater Slough.
-->

Spring/summer Chinook Salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014]. The DSR migrants leave the natal spawning area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal spawning areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migratory tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency.

The Deadwater Slough is in a reach of the Salmon River that is believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, this reach is part of the migratory pathway for juvenile DSR and NRR emigrants from all eight extant populations. The slough lacks hydrological and structural features (i.e., a homogenous channel with fine substrate and little cover) that can provide essential refuge from predation. As a result, predation on juvenile Chinook salmon proximal to the Deadwater Slough has been cited as a concern for the Upper Salmon River MPG, impacting DSR migrants in the fall and NRR emigrants during the spring.

@@

```{r}
# pull in SAR numbers
sar = read_csv(here("analysis/data/raw_data", 
                    "WSSCh_sar_4See.csv")) %>%
  summarise(across(c(medsar, sdsar),
                   median)) %>%
  summarise(across(c(medsar, sdsar),
               ~ . / 100))

adult_impact = fall_only_no_growth %>%
  filter(cum_cons_fish_g == max(cum_cons_fish_g)) %>%
  select(cum_cons_fish_g) %>%
  mutate(n_chnk_smolts = cum_cons_fish_g / 12) %>%
  bind_cols(all_est %>%
              select(event_name, 
                     model,
                     N,
                     Lci,
                     Uci)) %>%
  mutate(across(c(smolt_N = N, 
                  smolt_Lci = Lci, 
                  smolt_Uci = Uci),
                ~ . * n_chnk_smolts)) %>%
  bind_cols(sar) %>%
  mutate(adult_N = smolt_N * medsar,
         adult_Lci = smolt_Lci * qnorm(0.025, medsar, sdsar),
         adult_Uci = smolt_Uci * qnorm(0.975, medsar, sdsar))
```

To estimate the number of possible Chinook salmon that could be consumed, we divided the cumulative fish consumption in grams for an individual pikeminnow by the average weight of a Chinook emigrant to come up with a presmolt equivalency and then multiplied this by our abundance estimate of northern pikeminnow in Deadwater Slough, resulting in an estimate of the equivalent total number of Chinook salmon presmolts consumed. Finally, we estimated the number of additional adult Chinook salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough reach was reduced or eliminated. We multiplied the total estimated consumed juvenile Chinook salmon by an estimate of Granite-to-Granite SARs [@McCann2019].

@@

<!-- Currently only using one bioenergetics model run. Could incorporate more bioenergetics models, but if so then maybe consider focusing on a single abundance estimator (e.g. Schnabel) to avoid too many combinations in the results -->

@@

<!--
* Total population estimates of Pike Minnow. Chart or table, but likely a chart.
* What is that equate to as a density? e.g., how many pikeminnow per 10 square meters? Is that a reasonable number?
* To calculate density, need the area of the extent surveyed each year. Was that area consistent year-to-year?
-->

Abundance estimates for northern pikeminnow during the fall sampling events ranged from `r prettyNum(min(all_est$N), big.mark = ",")` to `r prettyNum(max(all_est$N), big.mark = ",")`. The multiple census estimators were consistently larger than the single census estimators, although the size of the 95% confidence intervals were more varied.

PSD for northern pikeminnow across all three years was 42.26% or 42.26% of fish were over the quality size classification 400 mm TL we denoted in the Methods, demonstrating a high percentage of above average size classes present within Deadwater Slough.

@@

```{r adult-imp-p}
adult_impact %>%
  mutate(stage = "Smolt") %>%
  select(event_name, model,
         stage,
         starts_with("smolt")) %>%
  rename_with(.fn = ~ str_remove(., "smolt_"),
              .cols = starts_with("smolt")) %>%
  bind_rows(adult_impact %>%
              mutate(stage = "Adult Equivalent") %>%
              select(event_name, model,
                     stage,
                     starts_with("adult")) %>%
              rename_with(.fn = ~ str_remove(., "adult_"),
                          .cols = starts_with("adult"))) %>%
  filter(str_detect(model, "Schuma", negate = T)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.5) +
  geom_point(size = 4) +
  facet_grid(stage ~ event_name,
             scales = "free_y") +
  scale_color_brewer(palette = "Set1",
                     name = "Model") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(x = "N. Pikeminnow Abundance Model")
```
